The Ranked Nearest Neighbour Interchange space

### 2019-06-26 paper.tex
minor

@@

The following Algorithm~\ref{alg:find_path} which computes paths between pairs of trees in $\rnni$ is a central tool for our investigation of shortest paths in $\rnni$.

@@

\todo{figure of an example?}

@@

Following the observations above, only the moves for the first three positions of a sequence with Bubble Sort need to be adapted for the computation of shortest caterpillar paths.
Let us consider the first three elements $a,b,c$ which appear in this order in a sequence representing a caterpillar tree on a caterpillar path between trees $T$ and $R$.
If the rank of $parent(c)$ is higher than the ranks of $parent(a)$ and $parent(b)$ in $R$, there are no swaps needed for these taxa.
If just one of the two taxa $a$ and $b$ is on the other side of $c$ in $R$, this one swaps position with $c$ in the sequence.
But if both $a$ and $b$ are on the other side of $c$ in $R$, the one that has the parent with highest rank of all these three taxa in $R$ swaps with $c$ at first.
This makes sure that there is no additional swap of $a$ and $b$ needed on the path to $R$.

@@

In~\cite{Gavryushkin2017} upper and lower bound for the diameter of $\rnni$ are given.
Following the results of section~\ref{section:caterpillar_convex} one can directly improve this result by giving the exact diameter of the $\rnni$ graph.

@@

Since the counterexample in Figure~\ref{splitthm_counterexample} clearly shows that the version of the Split Theorem stated in~\cite{Gavryushkin2017} does not hold, we will now claim an alternative to this conjecture, the \emph{Cluster Theorem}.
Within the Cluster Theorem we are considering clusters instead of splits.
This is motivated by the fact that rooted phylogenetic trees can uniquely be represented by sets of clusters~\ref{steel2016}, but not by sets of splits as they cannot define the position of the root.
Moreover, a ranked tree can be uniquely determined by ordering the set of clusters increasingly according to the ranks of the internal nodes inducing those clusters.

@@

\begin{lemma}
    Let $T$ and $R$ be caterpillar trees.
    It is $d_c(T,R) = \Delta(\rnniu)$ if, and only if, $T$ and $R$ share no induced triplet.
    \todo{define induced triplets}
\end{lemma}

@@

\begin{conjecture}
    Let $T$ and $R$ be trees on $n$ taxa.
    It is $d(T,R) = \Delta(\rnniu)$ if, and only if, $R$ can be produced from $T$ by applying Algorithm~\ref{alg:max_dist_tree}.
\end{conjecture}

According to Algorithm~\ref{alg:max_dist_tree}, the number of trees with maximum distance to a given tree $T$ depends on the number of cherries of $T$ and the ranks of their internal nodes.
For caterpillar trees it is $(n-1)!$.
The number of caterpillar trees with distance $\Delta(\rnniu)$ from a given caterpillar tree is $2^{n-2}$.

>>>>

%Figure of ranked tree in introduction

@@

Algorithm~\ref{alg:find_path} computes paths between pairs of trees in $\rnni$ and will be a central tool for our investigation of shortest paths in $\rnni$.

@@

\todo{figure of an example? Brief explanation what algorithm does?}

@@

Following the observations above, only the moves for the first three positions of a sequence with Bubble Sort need to be adapted for the computation of shortest caterpillar paths.
Let us consider the first three elements $a,b,c$ which appear in this order in a sequence representing a caterpillar tree on a caterpillar path between trees $T$ and $R$.
If the rank of $parent(c)$ is higher than the ranks of $parent(a)$ and $parent(b)$ in $R$, there are no swaps needed for these taxa.
If just one of the two taxa $a$ and $b$ is on the other side of $c$ in $R$, this one swaps position with $c$ in the sequence.
But if both $a$ and $b$ are on the other side of $c$ in $R$, the one that has the parent with highest rank of all these three taxa in $R$ swaps with $c$ at first.
This makes sure that there is no additional swap of $a$ and $b$ needed on the path to $R$.
\todo{Pseudo-Code for algorithm?}

@@

Following the results of section~\ref{section:caterpillar_convex} one can directly improve this result by giving the exact diameter of the $\rnni$ graph.

@@

Since the counterexample in Figure~\ref{splitthm_counterexample} clearly shows that the version of the Split Theorem stated in~\cite{Gavryushkin2017} does not hold, we will now claim an alternative to this conjecture, the \emph{Cluster Theorem}.
Within the Cluster Theorem we are considering clusters instead of splits.
This is motivated by the fact that rooted phylogenetic trees can uniquely be represented by sets of clusters~\cite{steel2016}, but not by sets of splits as they cannot define the position of the root.
Moreover, a ranked tree can be uniquely determined by ordering the set of clusters increasingly according to the ranks of the internal nodes inducing those clusters.

@@

\begin{lemma}
    Let $T$ and $R$ be caterpillar trees.
    It is $d_c(T,R) = \Delta(\rnni)$ if, and only if, $T$ and $R$ share no induced triplet.
    \todo{define induced triplets}
\end{lemma}

@@

\begin{conjecture}
    Let $T$ and $R$ be trees on $n$ taxa.
    It is $d(T,R) = \Delta(\rnni)$ if, and only if, $R$ can be produced from $T$ by applying Algorithm~\ref{alg:max_dist_tree}.
\end{conjecture}

If this conjecture is true, we have that $\Delta(\rnni) = rad(\rnni)$.

According to Algorithm~\ref{alg:max_dist_tree}, the number of trees with maximum distance to a given tree $T$ depends on the number of cherries of $T$ and the ranks of their internal nodes.
For caterpillar trees it is $(n-1)!$.
The number of caterpillar trees with distance $\Delta(\rnni)$ from a given caterpillar tree is $2^{n-2}$.
